# Question 1: Data Inspection, Cleaning and Visualisation

## Rationale

The four datasets were loaded separately because they contain different types
of FIFA World Cup information. Relative file paths were used so that the code
can run from the GitHub project on another computer.

The missing-value symbol `"?"` was replaced with `NA`, which is R's standard
representation of missing data. The variables `Result`, `Stage`, `Country`, and
`ExtraTime` were converted to factors because they represent categories.
`HomePenalty` and `AwayPenalty` were converted to numeric variables, and their
missing values were replaced with zero as required.

A bar chart was selected to display `HomePenalty` because penalty goals are
discrete count values such as 0, 1, 2, 3, 4, and 5. A bar chart presents each
exact value separately and makes the frequency of each penalty total easy to
compare. Only matches with `PenaltyShootout == 1` were included because the
question specifically asks about matches that involved a penalty shootout.

## Load the required package

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
```

## Load the datasets

```{r load-data}
# Load the four FIFA World Cup datasets from the project data folder.

Tournaments <- read.csv(
  here("data", "Tournaments.csv"),
  stringsAsFactors = FALSE
)

Teams <- read.csv(
  here("data", "Teams.csv"),
  stringsAsFactors = FALSE
)

Stadiums <- read.csv(
  here("data", "Stadiums.csv"),
  stringsAsFactors = FALSE
)

Matches <- read.csv(
  here("data", "Matches.csv"),
  stringsAsFactors = FALSE
)
```

## Inspect the data structure

```{r inspect-structure}
# Inspect the dimensions, variable names, and original data types.

str(Tournaments)
str(Teams)
str(Stadiums)
str(Matches)
```

## Present the data

```{r present-data}
# Display the first six observations from each dataset.

head(Tournaments)
head(Teams)
head(Stadiums)
head(Matches)
```

`str()` is used to inspect the structure and variable types, while `head()`
presents a small sample of each dataset without printing every observation.

## Replace `"?"` with `NA`

```{r replace-missing-symbols}
# Replace every question mark with R's standard missing value.

Tournaments[Tournaments == "?"] <- NA
Teams[Teams == "?"] <- NA
Stadiums[Stadiums == "?"] <- NA
Matches[Matches == "?"] <- NA
```

This method searches each data frame and changes all occurrences of `"?"` to
`NA`.

## Convert categorical variables to factors

```{r convert-factors}
# Convert the required categorical variables to factors.

Matches$Result <- factor(Matches$Result)
Matches$Stage <- factor(Matches$Stage)
Stadiums$Country <- factor(Stadiums$Country)
Matches$ExtraTime <- factor(Matches$ExtraTime)
```

The actual column name in the dataset is `ExtraTime`. R is case-sensitive, so
writing `Extratime` would create or refer to a different variable.

## Convert penalty variables to numeric and replace missing values

```{r clean-penalty-variables}
# Convert the two penalty variables from character to numeric format.

Matches$HomePenalty <- as.numeric(Matches$HomePenalty)
Matches$AwayPenalty <- as.numeric(Matches$AwayPenalty)

# Replace missing penalty values with zero.

Matches$HomePenalty[is.na(Matches$HomePenalty)] <- 0
Matches$AwayPenalty[is.na(Matches$AwayPenalty)] <- 0
```

The missing values are replaced after numeric conversion because the assignment
specifically requires missing `HomePenalty` and `AwayPenalty` values to become
zero.

## Inspect the cleaned data

```{r inspect-cleaned-data}
# Confirm that the required variables now have the correct data types.

str(Matches)
str(Stadiums)

# Present a summary of the cleaned Matches dataset.

summary(Matches)
```

## Visualise the distribution of HomePenalty

```{r home-penalty-plot, fig.width=8, fig.height=5}
# Use only matches that involved a penalty shootout in this chart.

ggplot(
  data = subset(Matches, PenaltyShootout == 1),
  aes(x = factor(HomePenalty))
) +
  geom_bar(
    fill = "steelblue",
    colour = "black"
  ) +
  labs(
    title = "Distribution of Home Penalty Goals in Penalty Shootouts",
    x = "Number of home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

The expression:

```r
subset(Matches, PenaltyShootout == 1)
```

does not create a separate filtering analysis. It simply ensures that the chart
uses only the matches requested by the question.

## Justification of the chart

A bar chart is appropriate because `HomePenalty` is a discrete numerical
variable with a small number of possible whole-number values. Each bar
represents one exact number of penalty goals, and the height represents how many
matches recorded that value.

A histogram was not selected because histograms are generally more useful for
continuous numerical variables that need to be divided into intervals. Grouping
these penalty totals into intervals would make the exact frequencies less clear.

## Interpretation

The chart contains 35 matches that involved a penalty shootout. Three home-team
penalty goals were the most common outcome, occurring in 12 matches. Four goals
were the second most common, occurring in 10 matches. Home teams recorded zero
goals in five matches, five goals in four matches, one goal in three matches,
and two goals in one match.

Therefore, the distribution is concentrated mainly around three and four
home-team penalty goals. Very low penalty totals occurred less frequently.
However, this is a descriptive analysis only. It does not demonstrate that
being recorded as the home team caused a team to score a particular number of
penalties.

## Code testing

```{r code-testing}
# Confirm that the required categorical variables are factors.

stopifnot(is.factor(Matches$Result))
stopifnot(is.factor(Matches$Stage))
stopifnot(is.factor(Stadiums$Country))
stopifnot(is.factor(Matches$ExtraTime))

# Confirm that the penalty variables are numeric.

stopifnot(is.numeric(Matches$HomePenalty))
stopifnot(is.numeric(Matches$AwayPenalty))

# Confirm that the penalty variables have no missing values.

stopifnot(!any(is.na(Matches$HomePenalty)))
stopifnot(!any(is.na(Matches$AwayPenalty)))

# Confirm that the dataset contains penalty-shootout matches.

stopifnot(sum(Matches$PenaltyShootout == 1) == 35)

cat("All Question 1 tests passed successfully.")
```