---
output:
  pdf_document: default
  html_document: default
---
# Question 1: Data Inspection, Cleaning and Visualisation

## Rationale

The four datasets were loaded separately because they contain different types
of FIFA World Cup information. Relative file paths were used so that the code
can run from the GitHub project on another computer.

The missing-value symbol `"?"` was replaced with `NA`, which is R's standard
representation of missing data. The variables `Result`, `Stage`, `Country`, and
`ExtraTime` were converted to factors because they represent categories.
`HomePenalty` and `AwayPenalty` were converted to numeric variables, and their
missing values were replaced with zero as required.

A bar chart was selected to display `HomePenalty` because penalty goals are
discrete count values such as 0, 1, 2, 3, 4, and 5. A bar chart presents each
exact value separately and makes the frequency of each penalty total easy to
compare. Only matches with `PenaltyShootout == 1` were included because the
question specifically asks about matches that involved a penalty shootout.

## Load the required package

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
```

## Load the datasets

```{r load-data}
# Load the four FIFA World Cup datasets from the project data folder.

 #Exercise 1
  Tournaments <- read.csv("E:/A2 data/Tournaments.csv")
  Teams <- read.csv("E:/A2 data/Teams.csv")
  Stadiums <- read.csv("E:/A2 data/Stadiums.csv")
  Matches <- read.csv("E:/A2 data/Matches.csv")

```

## Inspect the data structure

```{r inspect-structure}
# Inspect the dimensions, variable names, and original data types.

str(Tournaments)
str(Teams)
str(Stadiums)
str(Matches)
```

## Present the data

```{r present-data}
# Display the first six observations from each dataset.

head(Tournaments)
head(Teams)
head(Stadiums)
head(Matches)
```

`str()` is used to inspect the structure and variable types, while `head()`
presents a small sample of each dataset without printing every observation.

## Replace `"?"` with `NA`

```{r replace-missing-symbols}
# Replace every question mark with R's standard missing value.

Tournaments[Tournaments == "?"] <- NA
Teams[Teams == "?"] <- NA
Stadiums[Stadiums == "?"] <- NA
Matches[Matches == "?"] <- NA
```

This method searches each data frame and changes all occurrences of `"?"` to
`NA`.

## Convert categorical variables to factors

```{r convert-factors}
# Convert the required categorical variables to factors.

Matches$Result <-as.factor(Matches$Result)
Matches$Stage <- as.factor(Matches$Stage)
Stadiums$Country <- as.factor(Stadiums$Country)
Matches$ExtraTime <- as.factor(Matches$ExtraTime)
```

The actual column name in the dataset is `ExtraTime`. R is case-sensitive, so
writing `Extratime` would create or refer to a different variable.

## Convert penalty variables to numeric and replace missing values

```{r clean-penalty-variables}
# Convert the two penalty variables from character to numeric format.

Matches$HomePenalty <- as.numeric(Matches$HomePenalty)
Matches$AwayPenalty <- as.numeric(Matches$AwayPenalty)

# Replace missing penalty values with zero.

Matches$HomePenalty[is.na(Matches$HomePenalty)] <- 0
Matches$AwayPenalty[is.na(Matches$AwayPenalty)] <- 0
```

The missing values are replaced after numeric conversion because the assignment
specifically requires missing `HomePenalty` and `AwayPenalty` values to become
zero.

## Inspect the cleaned data

```{r inspect-cleaned-data}
# Confirm that the required variables now have the correct data types.

str(Matches)
str(Stadiums)

# Present a summary of the cleaned Matches dataset.

summary(Matches)
```

## Identify matches that involved a penalty shootout

```{r subset-penalty-shootouts}
# Keep only matches that involved a penalty shootout.
# In the dataset, 1 means that a penalty shootout occurred.

PenaltyMatches <- subset(
  Matches,
  PenaltyShootout == 1
)

# Inspect the filtered data.

head(PenaltyMatches)
nrow(PenaltyMatches)
```

The `subset()` function was used to select only observations where
`PenaltyShootout` equals 1. This ensures that the following visualisation does
not include ordinary matches that did not proceed to a penalty shootout.

## Visualise the distribution of HomePenalty

```{r home-penalty-visualisation, fig.width=8, fig.height=5}
ggplot(
  data = PenaltyMatches,
  aes(x = factor(HomePenalty))
) +
  geom_bar(
    fill = "steelblue",
    colour = "black"
  ) +
  labs(
    title = "Distribution of Home Penalty Goals in Penalty Shootouts",
    x = "Number of home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Justification of chart choice

A bar chart was selected because `HomePenalty` is a discrete numerical variable.
The possible values are whole-number counts such as 0, 1, 2, 3, 4, and 5.
Each bar therefore represents one exact number of home-team penalty goals, and
the height of the bar shows how many matches recorded that value.

Although `HomePenalty` was previously converted to numeric format, it is
temporarily written as `factor(HomePenalty)` inside `aes()`. This tells
`ggplot2` to display each possible penalty total as a separate category. It
does not change the stored data type of `Matches$HomePenalty`.

A histogram was not selected because a histogram divides numerical values into
intervals. For this variable, the exact frequencies of individual whole-number
penalty totals are more meaningful than grouped intervals.

## Interpretation

The chart shows that three home-team penalty goals were the most frequent
result, occurring in 12 penalty-shootout matches. Four goals were the second
most frequent result, occurring in 10 matches.

Zero goals occurred in five matches, five goals occurred in four matches, one
goal occurred in three matches, and two goals occurred in only one match.
Therefore, the distribution is mainly concentrated around three and four
home-team penalty goals.

This is a descriptive result. It shows the pattern present in the dataset, but
it does not prove that being listed as the home team caused a team to score a
particular number of penalties.

## Code testing

```{r code-testing}
# Confirm that the required categorical variables are factors.

stopifnot(is.factor(Matches$Result))
stopifnot(is.factor(Matches$Stage))
stopifnot(is.factor(Stadiums$Country))
stopifnot(is.factor(Matches$ExtraTime))

# Confirm that the penalty variables are numeric.

stopifnot(is.numeric(Matches$HomePenalty))
stopifnot(is.numeric(Matches$AwayPenalty))

# Confirm that the penalty variables have no missing values.

stopifnot(!any(is.na(Matches$HomePenalty)))
stopifnot(!any(is.na(Matches$AwayPenalty)))

# Confirm that the dataset contains penalty-shootout matches.

stopifnot(sum(Matches$PenaltyShootout == 1) == 35)

cat("All Question 1 tests passed successfully.")
```
# Question 2: Distribution of HomeTeamScore

## Rationale

The `Matches` dataset contains `HomeTeamScore`, `AwayTeamScore`, and `Stage`,
while `TournamentName` is stored in the `Tournaments` dataset. Therefore, the
two datasets were combined using their common variable, `TournamentID`.
A left-style merge was used to retain every observation from the Matches
dataset.

Histograms were selected because the task specifically requires histograms to
analyse the distribution of `HomeTeamScore`. A bin width of one was used because
football scores are recorded as whole numbers. Faceting was used to display a
separate histogram for each competition stage, tournament, and away-team score
group.

The `AwayTeamScore` values were divided into the three groups specified in the
task: 0–1 goals, 2–3 goals, and 4 or more goals. The `cut()` function was used
because it provides clear numerical boundaries for these groups.

## Combine Matches and Tournaments

```{r q2-combine-data}
# Add TournamentName to the Matches dataset using TournamentID.

Matches2 <- merge(
  Matches,
  Tournaments[, c("TournamentID", "TournamentName")],
  by = "TournamentID",
  all.x = TRUE,
  sort = FALSE
)

# Present the first six rows of the combined dataset.

head(Matches2)
```

`all.x = TRUE` keeps all observations from the `Matches` dataset, even if a
matching tournament record is not found.

## Distribution of HomeTeamScore by Stage

```{r q2-stage-histogram, fig.width=10, fig.height=9}
ggplot(
  Matches2,
  aes(x = HomeTeamScore)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = 0:max(Matches2$HomeTeamScore, na.rm = TRUE)
  ) +
  facet_wrap(
    ~ Stage,
    ncol = 2,
    scales = "free_y"
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Competition Stage",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

`scales = "free_y"` allows each panel to use a suitable vertical scale.
This makes the distribution visible for stages containing fewer matches.
Therefore, the shapes of the distributions can be compared, but the absolute
bar heights should be compared carefully because the vertical scales may differ.

## Interpretation by Stage

Across most competition stages, home-team scores are concentrated between zero
and two goals. In the group stage, one goal is the most frequent result, while
zero and two goals are also common. The round of 16 and quarter-finals show a
similar concentration around one and two goals.

Finals show a somewhat different pattern, with three and four home-team goals
appearing relatively often. However, there are only 21 final matches, so this
pattern is based on a much smaller sample than the group-stage distribution.
A small number of unusually high scores also create long right tails in several
stages.

Overall, most stage distributions are right-skewed: lower scores are common,
while very high home-team scores occur only occasionally.

## Distribution of HomeTeamScore by TournamentName

```{r q2-tournament-histogram, fig.width=12, fig.height=15}
ggplot(
  Matches2,
  aes(x = HomeTeamScore)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = 0:max(Matches2$HomeTeamScore, na.rm = TRUE)
  ) +
  facet_wrap(
    ~ TournamentName,
    ncol = 4,
    scales = "free_y"
  ) +
  labs(
    title = "Distribution of Home-Team Scores by FIFA World Cup",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 7),
    axis.text.x = element_text(size = 7)
  )
```

## Interpretation by Tournament

The earlier tournaments generally show wider score distributions and more
high-scoring home teams. This is particularly visible in tournaments such as
1930, 1938, 1950, and 1954. The 1954 tournament has a distribution concentrated
further to the right, with four home-team goals occurring frequently.

From the 1970s onward, the distributions become more concentrated around zero,
one, and two home-team goals. Modern tournaments such as 1990, 2006, 2010, and
2014 show comparatively fewer high home-team scores.

This suggests that the distribution of home-team goals changed across
tournaments, with earlier World Cups generally having wider and higher-scoring
distributions. However, tournament sizes and competition formats changed over
time, so the histograms describe trends but do not identify their causes.

## Create AwayTeamScore groups

```{r q2-away-score-groups}
# Divide AwayTeamScore into the three groups required by the task.

Matches2$AwayScoreGroup <- cut(
  Matches2$AwayTeamScore,
  breaks = c(-1, 1, 3, Inf),
  labels = c(
    "0-1 goals",
    "2-3 goals",
    "4 or more goals"
  ),
  include.lowest = TRUE
)

# Check the new variable.

str(Matches2$AwayScoreGroup)
```

The boundaries create the following groups:

- values from 0 to 1 become `"0-1 goals"`;
- values from 2 to 3 become `"2-3 goals"`;
- values of 4 or more become `"4 or more goals"`.

## Distribution of HomeTeamScore by AwayTeamScore group

```{r q2-away-group-histogram, fig.width=10, fig.height=5}
ggplot(
  Matches2,
  aes(x = HomeTeamScore)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = 0:max(Matches2$HomeTeamScore, na.rm = TRUE)
  ) +
  facet_wrap(
    ~ AwayScoreGroup,
    nrow = 1,
    scales = "free_y"
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Away-Team Score Group",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Interpretation by AwayTeamScore group

For matches where the away team scored zero or one goal, home-team scores are
mainly concentrated between zero and two goals, with one goal being the most
common result. This group contains most of the observations.

For matches where the away team scored two or three goals, home-team scores
remain concentrated around zero, one, and two goals. The distribution is
slightly more spread out, although high home-team scores are still uncommon.

The group where the away team scored four or more goals contains relatively few
matches. Home-team scores in this group are also generally low, with one goal
being the most common result. Because this group has a much smaller number of
observations, its pattern should be interpreted cautiously.

Overall, the shapes of the three distributions are broadly similar. In every
group, low home-team scores occur more frequently than very high home-team
scores. There is no strong visual evidence that higher away-team score groups
are associated with a major shift in the typical home-team score.

## Code testing

```{r q2-code-testing}
# The merge should preserve every row from Matches.

stopifnot(nrow(Matches2) == nrow(Matches))

# Every match should receive a TournamentName.

stopifnot(!any(is.na(Matches2$TournamentName)))

# AwayScoreGroup should be a factor created by cut().

stopifnot(is.factor(Matches2$AwayScoreGroup))

# Every match should be assigned to one of the three groups.

stopifnot(!any(is.na(Matches2$AwayScoreGroup)))

# Test the boundaries of each AwayTeamScore group.

stopifnot(
  all(
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "0-1 goals"
    ] <= 1
  )
)

stopifnot(
  all(
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "2-3 goals"
    ] >= 2 &
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "2-3 goals"
    ] <= 3
  )
)

stopifnot(
  all(
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "4 or more goals"
    ] >= 4
  )
)

cat("All Question 2 tests passed successfully.")
```

If all tests are correct, the final output will display:

```text
All Question 2 tests passed successfully.
```
#Question 3: National Teams in Penalty Shootouts

##  Rationale 
The analysis first filters the Matches dataset to retain only matches that were
decided by a penalty shootout. Home-team and away-team participation are counted
separately because they are stored in different variables and the task requires
two separate rankings.

The `table()` function is used to count how many times each Team ID appears.
The resulting counts are then joined with the Teams dataset using `merge()` so
that each Team ID can be displayed using its official TeamName and TeamCode.
Finally, the results are sorted from highest to lowest and the first five teams
are selected.
##Filter out the matches decided by a penalty shootout
```{r q3-filter-penalty-matches}
Q3penalty_matches <- subset(
  Matches, 
  PenaltyShootout == 1 
)
#Check the number of penalty-shootour matches
nrow(Q3penalty_matches)
#Check the accuracy of filtered data 
table(Q3penalty_matches$PenaltyShootout)
```
##Top 5 home-teams that have most frequently participated in these matches
```{r q3-count-home-teams}
# Count how many penalty shootouts each home team participated in
HomeTeamCounts <- as.data.frame(
  table(Q3penalty_matches$HomeTeamID)
)

# Rename the columns
names(HomeTeamCounts) <- c(
  "TeamID",
  "NumberOfMatches"
)

# Make sure TeamID has the same data type in both datasets
HomeTeamCounts$TeamID <- as.character(HomeTeamCounts$TeamID)
Teams$TeamID <- as.character(Teams$TeamID)

# Add TeamName and TeamCode using TeamID
HomeTeamResults <- merge(
  HomeTeamCounts,
  Teams,
  by = "TeamID"
)

# Sort from highest to lowest
HomeTeamResults <- HomeTeamResults[
  order(-HomeTeamResults$NumberOfMatches),
]

# Keep the required columns
HomeTeamResults <- HomeTeamResults[
  ,
  c("TeamName", "TeamCode", "NumberOfMatches")
]

# Select the top five teams
Top5HomeTeams <- head(HomeTeamResults, 5)

Top5HomeTeams
```
##Top 5 away-teams that have most frequently participated in these matches
```{r q3-count-away-teams}
AwayTeamCounts <- as.data.frame(
  table(Q3penalty_matches$AwayTeamID)
)

names(AwayTeamCounts) <- c(
  "TeamID",
  "NumberOfMatches"
)

AwayTeamCounts$TeamID <- as.character(AwayTeamCounts$TeamID)

AwayTeamResults <- merge(
  AwayTeamCounts,
  Teams,
  by = "TeamID"
)

AwayTeamResults <- AwayTeamResults[
  order(-AwayTeamResults$NumberOfMatches),
]

AwayTeamResults <- AwayTeamResults[
  ,
  c("TeamName", "TeamCode", "NumberOfMatches")
]

Top5AwayTeams <- head(AwayTeamResults, 5)

Top5AwayTeams
```

