---
title: "AP Assignment Report"
author: "Le Van Vinh Phat"
date: "24 July 2026"
output: pdf_document
---
\begin{center}

{\LARGE\textbf{AP Assignment Report}}\\[0.8cm]

{\large\textbf{COMP1013 Analytics Programming}}\\[1cm]

{\large Le Van Vinh Phat}\\[0.3cm]

{\large Student Number: 22202375}\\[0.3cm]

{\large 24 July 2026}\\[0.5cm]

\end{center}

\vspace{0.7cm}
## Declaration

By including this statement, I, the author of this work, verify that:
- I hold a copy of this assignment that I can produce if the original is lost or damaged.

- I hereby certify that no part of this assignment/product has been copied from any other student's work or from any other source except where due acknowledgement is made in the assignment.

- No part of this assignment/product has been written/produced for me by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.

- I am aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism, which may retain a copy on its database for future plagiarism checking.

- I hereby certify that I have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

\newpage

# Question 1: Data Inspection, Cleaning and Visualisation
# Question 1: Data Inspection, Cleaning and Visualisation

## Rationale

The four data sets are imported independently due to the different types of data. Access to World Cup information. Relative project paths are used to ensure that the analysis can be run after downloading repository to another computer. Question Just like a number, a missing value is represented as 'NA', which is R's default representation for. missing data.
The factors is applied to the following variables:The following variables are converted to factors: HomePenalty is not a continuous measurement, they are categories. 'AwayPenalty' are converted to numeric variables so that they can be analysed Correctly, any missing data are filled with zeros as per task requirements.
The penalty shootout matches are chosen using 'subset()' and then plotted. A The choice of bar chart is chosen because 'HomePenalty' is a count variable that takes only discrete values and has a Few Integer values. exact frequency of each bar in separate figure Penalties are far more easy to identify with each interval in the grouped histogram than with each interval in the drawn histogram.


## Load the required package

```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(here)
```

## Load the datasets

```{r load-data}
# Load the four FIFA World Cup dataset from the project data folder.

 #Exercise 1
  Tournaments <- read.csv("E:/A2 data/Tournaments.csv")
  Teams <- read.csv("E:/A2 data/Teams.csv")
  Stadiums <- read.csv("E:/A2 data/Stadiums.csv")
  Matches <- read.csv("E:/A2 data/Matches.csv")

```

## Inspect the data structure

```{r inspect-structure}
# Inspect the dimensions, variable names, and original data types.

str(Tournaments)
str(Teams)
str(Stadiums)
str(Matches)
```

## Present the data

```{r present-data}
# Display the first six observations from each dataset.

head(Tournaments)
head(Teams)
head(Stadiums)
head(Matches)
```

`str()` is used to inspect the structure and variable types, while `head()`
presents a small sample of each dataset without printing every observation.

## Replace `"?"` with `NA`

```{r replace-missing-symbols}
# Replace every question mark with R's standard missing value.

Tournaments[Tournaments == "?"] <- NA
Teams[Teams == "?"] <- NA
Stadiums[Stadiums == "?"] <- NA
Matches[Matches == "?"] <- NA
```

This method searches each data frame and changes all occurrences of `"?"` to
`NA`.

## Convert categorical variables to factors

```{r convert-factors}
# Convert the required categorical variables to factors.

Matches$Result <-as.factor(Matches$Result)
Matches$Stage <- as.factor(Matches$Stage)
Stadiums$Country <- as.factor(Stadiums$Country)
Matches$ExtraTime <- as.factor(Matches$ExtraTime)
```

The actual column name in the dataset is `ExtraTime`. R is case-sensitive, so
writing `Extratime` would create or refer to a different variable.

## Convert penalty variables to numeric and replace missing values

```{r clean-penalty-variables}
# Convert the two penalty variables from character to numeric format.

Matches$HomePenalty <- as.numeric(Matches$HomePenalty)
Matches$AwayPenalty <- as.numeric(Matches$AwayPenalty)

# Replace missing penalty values with zero.

Matches$HomePenalty[is.na(Matches$HomePenalty)] <- 0
Matches$AwayPenalty[is.na(Matches$AwayPenalty)] <- 0
```

The missing values are replaced after numeric conversion because the assignment
specifically requires missing `HomePenalty` and `AwayPenalty` values to become
zero.

## Inspect the cleaned data

```{r inspect-cleaned-data}
# Confirm that the required variables now have the correct data types.

str(Matches)
str(Stadiums)

# Present a summary of the cleaned Matches dataset.

summary(Matches)
```

## Identify matches that involved a penalty shootout

```{r subset-penalty-shootouts}
# Keep only matches that involved a penalty shootout.
# In the dataset, 1 means that a penalty shootout occurred.

PenaltyMatches <- subset(
  Matches,
  PenaltyShootout == 1
)

# Inspect the filtered data.

head(PenaltyMatches)
nrow(PenaltyMatches)
```

The `subset()` function was used to select only observations where
`PenaltyShootout` equals 1. This ensures that the following visualisation does
not include ordinary matches that did not proceed to a penalty shootout.

## Visualise the distribution of HomePenalty

```{r home-penalty-visualisation, fig.width=8, fig.height=5}
ggplot(
  data = PenaltyMatches,
  aes(x = factor(HomePenalty))
) +
  geom_bar(
    fill = "steelblue",
    colour = "black"
  ) +
  labs(
    title = "Distribution of Home Penalty Goals in Penalty Shootouts",
    x = "Number of home-team penalty goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Justification of chart choice

Because of being a discrete numerical variable, the bar chart will be the most suitable graph to display ‘HomePenalty’. In this, each bar can represent one exact number of home-team penalty goals and the height of the bar shows the frequency of matches.

While ‘Matches$HomePenalty’ is permanently stored as a numeric value in the dataset, it is temporarily treated as categorical data by writing factor(HomePenalty) in the aes() function. This tells the plotting engine to separate every penalty count without permanently altering the underlying dataframe. 

Despite the unsuitable impact of histogram that divides numerical values into intervals, it can meet the demand of these variables that the frequencies of individuals whole-number penalty totals are more meaningful than grouped intervals. 

## Interpretation

The chart includes 35 matches that were decided by a penalty shootout. Three The most common, occurring in 12 games, is home-team penalty goal. Gave up four goals in 10 games. These values add up to 22 of the 35 observations, with the distribution being primarily centered around three. Four successful penalties, and.
Five matches with zero goals, 3 matches with one, 1 match with 2. The five goals he has scored in his last four games, matches. The zero category is valid since a team can shoot but not score points in an original shootout. There are no missing values of 'HomePenalty' in observations.
The distribution is skewed to the right with the majority of observations being towards the lower end. There are fewer matches at the other values; near three and four. Overall, three This data set contains the most common home team scores: four penalties. The chart is descriptive but it is not proof that the home-team A specific score of penalty was awarded for designation.


## Code testing

```{r code-testing}
# Confirm that the required categorical variables are factors.

stopifnot(is.factor(Matches$Result))
stopifnot(is.factor(Matches$Stage))
stopifnot(is.factor(Stadiums$Country))
stopifnot(is.factor(Matches$ExtraTime))

# Confirm that the penalty variables are numeric.

stopifnot(is.numeric(Matches$HomePenalty))
stopifnot(is.numeric(Matches$AwayPenalty))

# Confirm that the penalty variables have no missing values.

stopifnot(!any(is.na(Matches$HomePenalty)))
stopifnot(!any(is.na(Matches$AwayPenalty)))

# Confirm that the dataset contains penalty-shootout matches.

stopifnot(sum(Matches$PenaltyShootout == 1) == 35)

cat("All Question 1 tests passed successfully.")
```
# Question 2: Distribution of HomeTeamScore

## Rationale

The dataset 'Matches' includes the columns 'HomeTeamScore', 'AwayTeamScore' and 'Stage', while 'TournamentName' is stored in 'Tournaments'. Because of this, the datasets are therefore The tournament data was linked to the players' data by their shared 'TournamentID' so that each match could be analysed. by tournament name.
A) they are requested in the task, and B) they are appropriate to use for this type of data. they show how frequently different home-team scores occur. A bin width of one is suitable since the scores in football are whole number counts. The function 'facet_wrap()' will make a histogram for each stage, tournament, and Using away-score group, multiple distributions can be compared in a single analysis. figure.
AwayTeamScore is segmented into 0-1, 2-3 and 4+ goal groups with cut(), which does claret and repeatable numeric boundaries. The analysis is concerned with distribution shape, centre, spread and unusually high. To compare scores keeping in mind the difference in the number of matches played in each group.

## Combine Matches and Tournaments

```{r q2-combine-data}
# Add TournamentName to the Matches dataset using TournamentID.

Matches2 <- merge(
  Matches,
  Tournaments[, c("TournamentID", "TournamentName")],
  by = "TournamentID",
  all.x = TRUE,
  sort = FALSE
)

# Present the first six rows of the combined dataset.

head(Matches2)
```

`all.x = TRUE` keeps all observations from the `Matches` dataset, even if a
matching tournament record is not found.

## Distribution of HomeTeamScore by Stage

```{r q2-stage-histogram, fig.width=10, fig.height=9}
ggplot(
  Matches2,
  aes(x = HomeTeamScore)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = 0:max(Matches2$HomeTeamScore, na.rm = TRUE)
  ) +
  facet_wrap(
    ~ Stage,
    ncol = 2,
    scales = "free_y"
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Competition Stage",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

`scales = "free_y"` allows each panel to use a suitable vertical scale.
This makes the distribution visible for stages containing fewer matches.
Therefore, the shapes of the distributions can be compared, but the absolute
bar heights should be compared carefully because the vertical scales may differ.

## Interpretation by Stage

In most competitons, the home team scores fall around the average of zero. and two goals. This stage of the group stage has 676 matches with a median score of one goal, the most common outcome being one goal. There is a round of 16 in the event. A slightly higher centre with a mean of approximately 2.04 and a median of 2.
Medians are also two for a semi-final and other third place matches have a median of three goals. The last round is the highest round with the mean of the approximate value of It has a 3.83 rating, but includes just 6 matches. For this reason, there are only a few high-quality quizzes available. Its distribution can be significantly affected by its scores.
Several stages have long right tails as occasionally matches contain: High away team scores, even as high as 10 in the group stage. A general trend of lower scores is observed, except for the later knockout In the cases of some stages, higher centres are shown. The differences are to be interpreted are used with great care since the size of samples on the stages varies considerably and the charts do not Consider what factors, like a strong team or old tournament regulations, may have influenced the results.

## Distribution of HomeTeamScore by TournamentName

```{r q2-tournament-histogram, fig.width=12, fig.height=15}
ggplot(
  Matches2,
  aes(x = HomeTeamScore)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = 0:max(Matches2$HomeTeamScore, na.rm = TRUE)
  ) +
  facet_wrap(
    ~ TournamentName,
    ncol = 4,
    scales = "free_y"
  ) +
  labs(
    title = "Distribution of Home-Team Scores by FIFA World Cup",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 7),
    axis.text.x = element_text(size = 7)
  )
```

## Interpretation by Tournament

The histograms of the tournaments show that there were more World Cups with higher and More dispersed home-team score distributions than previous tournaments. The 1954 FIFA On average, World Cup has the highest amount of goals scored by the home team, around 4.19 goals, followed by 1938 at 3.39 and 1930 at 3.28. These tournaments are also included in the games several matches where the home team scored four or more goals.
That is contrasted to modern tournaments, which tend to focus on the 0-1 area, and two goals. The mean was approximately 1.23 in 1990, 1.19 in 2010, and 1.27 in 2014. Some recent competitions still have a few high-scoring isolated competitions. In matches, these observations are not the, but are instead part of long right tails. typical score.
As a whole, the pattern of scores indicates that home teams started to score less. More focussed through time. However, the analysis doesn't provide a cause as to why this occurred. The level of the tournament, the type of competition, the quality of the teams, the way the game is played and the defensive tactics have evolved over time, and the name of the tournament (TournamentName) has changed over time. represents multiple differences and not just one cause of that difference. 

## Create AwayTeamScore groups

```{r q2-away-score-groups}
# Divide AwayTeamScore into the three groups required by the task.

Matches2$AwayScoreGroup <- cut(
  Matches2$AwayTeamScore,
  breaks = c(-1, 1, 3, Inf),
  labels = c(
    "0-1 goals",
    "2-3 goals",
    "4 or more goals"
  ),
  include.lowest = TRUE
)

# Check the new variable.

str(Matches2$AwayScoreGroup)
```

The boundaries create the following groups:

- values from 0 to 1 become `"0-1 goals"`;
- values from 2 to 3 become `"2-3 goals"`;
- values of 4 or more become `"4 or more goals"`.

## Distribution of HomeTeamScore by AwayTeamScore group

```{r q2-away-group-histogram, fig.width=10, fig.height=5}
ggplot(
  Matches2,
  aes(x = HomeTeamScore)
) +
  geom_histogram(
    binwidth = 1,
    boundary = -0.5,
    fill = "steelblue",
    colour = "black"
  ) +
  scale_x_continuous(
    breaks = 0:max(Matches2$HomeTeamScore, na.rm = TRUE)
  ) +
  facet_wrap(
    ~ AwayScoreGroup,
    nrow = 1,
    scales = "free_y"
  ) +
  labs(
    title = "Distribution of Home-Team Scores by Away-Team Score Group",
    x = "Home-team goals",
    y = "Number of matches"
  ) +
  theme_minimal()
```

## Interpretation by AwayTeamScore group

Away-Team Score Group interpretation is available.Away-Team Score Group interpretation is available.
The score of the 3 'AwayTeamScore' groups is fairly similar at home. distributions. In each group, the goal scored by the home team is the most common value. This is also true of the median of the data.
The mean HomeTeamScore is ~1.77 when the away team scored. Any goal scored by the ball was equal to 1.75 when the ball scored two goals, and 1.73 when the ball scored three goals. scored 4 or more. The similar averages suggest that the centre of the This is because there are only minor differences in home-team distribution between the groups.
The biggest difference is the number of samples. The 0-1 group have 701 matches, 237 are in the 2–3 group and there are only 26 in the 4-or-more group. Consequently, The last one is less smooth, and more susceptible to match-ups.
Overall the charts do not offer much visual clue of a significant change in usual. The value of 'HomeTeamScore' gets higher as the value of 'AwayTeamScore' also gets higher. This is a description, a comparison, and the smallest highest scoring group restricts the force of any conclusion


## Code testing

```{r q2-code-testing}
# The merge should preserve every row from Matches.

stopifnot(nrow(Matches2) == nrow(Matches))

# Every match should receive a TournamentName.

stopifnot(!any(is.na(Matches2$TournamentName)))

# AwayScoreGroup should be a factor created by cut().

stopifnot(is.factor(Matches2$AwayScoreGroup))

# Every match should be assigned to one of the three groups.

stopifnot(!any(is.na(Matches2$AwayScoreGroup)))

# Test the boundaries of each AwayTeamScore group.

stopifnot(
  all(
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "0-1 goals"
    ] <= 1
  )
)

stopifnot(
  all(
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "2-3 goals"
    ] >= 2 &
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "2-3 goals"
    ] <= 3
  )
)

stopifnot(
  all(
    Matches2$AwayTeamScore[
      Matches2$AwayScoreGroup == "4 or more goals"
    ] >= 4
  )
)

cat("All Question 2 tests passed successfully.")
```

If all tests are correct, the final output will display:

```text
All Question 2 tests passed successfully.
```
#Question 3: National Teams in Penalty Shootouts

##  Rationale 
Question 3 first uses 'subset()' to select only those matches that 'PenaltyShootout' equals one. Home and Away teams are counted. Are stored separately (HomeTeamID/AwayTeamID) and There are two rankings for task.
The function table() is used to see the frequency of each TeamID value and as.data.frame() converts it to a dataframe. Convert the counts into a structure which can be sorted and joined. The counts are Merged to 'Teams' using 'TeamID' as the key as there are no unique keys to merge on in 'Teams' and therefore 'TeamName' and 'TeamCode' are not The data is stored in the 'Matches' data set.
'order()' is a function which sorts the teams from the best to the worst The page with the numbers 1, 2, 3, 4, and 5 will be returned by the query NumberOfMatches, and by head(..., 5). The final The only difference between the tables and the table models is that the tables only contain the 'TeamName', 'TeamCode', and 'NumberOfMatches' properties, as do the table models. required output. When using alphabetical ordering, it can be used as a constant rule when The number of participants in each team is the same.

##Filter out the matches decided by a penalty shootout
```{r q3-filter-penalty-matches}
Q3penalty_matches <- subset(
  Matches, 
  PenaltyShootout == 1 
)
#Check the number of penalty-shootour matches
nrow(Q3penalty_matches)
#Check the accuracy of filtered data 
table(Q3penalty_matches$PenaltyShootout)
```
##Top 5 home-teams that have most frequently participated in these matches
```{r q3-count-home-teams}
# Count how many penalty shootouts each home team participated in
HomeTeamCounts <- as.data.frame(
  table(Q3penalty_matches$HomeTeamID)
)

# Rename the columns
names(HomeTeamCounts) <- c(
  "TeamID",
  "NumberOfMatches"
)

# Make sure TeamID has the same data type in both datasets
HomeTeamCounts$TeamID <- as.character(HomeTeamCounts$TeamID)
Teams$TeamID <- as.character(Teams$TeamID)

# Add TeamName and TeamCode using TeamID
HomeTeamResults <- merge(
  HomeTeamCounts,
  Teams,
  by = "TeamID"
)

# Sort from highest to lowest
HomeTeamResults <- HomeTeamResults[
  order(-HomeTeamResults$NumberOfMatches),
]

# Keep the required columns
HomeTeamResults <- HomeTeamResults[
  ,
  c("TeamName", "TeamCode", "NumberOfMatches")
]

# Select the top five away teams
Top5HomeTeams <- head(HomeTeamResults, 5)

Top5HomeTeams
```
##Top 5 away-teams that have most frequently participated in these matches
```{r q3-count-away-teams}
AwayTeamCounts <- as.data.frame(
  table(Q3penalty_matches$AwayTeamID)
)

names(AwayTeamCounts) <- c(
  "TeamID",
  "NumberOfMatches"
)

AwayTeamCounts$TeamID <- as.character(AwayTeamCounts$TeamID)

AwayTeamResults <- merge(
  AwayTeamCounts,
  Teams,
  by = "TeamID"
)

AwayTeamResults <- AwayTeamResults[
  order(-AwayTeamResults$NumberOfMatches),
]

AwayTeamResults <- AwayTeamResults[
  ,
  c("TeamName", "TeamCode", "NumberOfMatches")
]
#Select the top five away-teams
Top5AwayTeams <- head(AwayTeamResults, 5)

Top5AwayTeams
```
## Discussion
There is a considerable difference between the ranking of the home team and the ranking of the away team. Argentina, Brazil, Home teams shared the lead with four penalty shootout with Spain. appearances each. 348, followed by the Netherlands and West Germany with three.Three after that, followed by the Netherlands and West Germany, with 348. appearances.
Fourth-placed France has five appearances, followed by Italy and Switzerland. There are three in Argentina and three in England. Italy and Croatia fill the away half.The away half is rounded off by Croatia and Italy. 5 players with 2 starts apiece.
Argentina is the only team that is in both top-5s, meaning: Teams' participation, whether at home or away, was high, particularly when not recorded. away side. France are the number one away team but not in the home top. Brazil and Spain have the opposite pattern as shown by the numbers five.
These are participation-based ranks and not success-based ranks. They aren't a way to show how A number of shootouts that each team had won. They may not be used as evidence of any kind either. There are not a lot of home and away labels to be identified, except the labels World Cup home and World Cup away are mostly used to indicate of home ground advantage. In a match, the two teams that are pitted against each other. In general, the two roles are separated and the Ranking and shows trends that otherwise wouldn't be observed in a cumulative total.

## Code testing

```{r q3-code-testing}
# Confirm that every filtered observation is a penalty-shootout match.

stopifnot(
  all(Q3penalty_matches$PenaltyShootout == 1)
)

# Confirm that the filtered dataset is not empty.

stopifnot(
  nrow(Q3penalty_matches) > 0
)

# Every shootout match must contribute one home-team observation.

stopifnot(
  sum(HomeTeamCounts$NumberOfMatches) ==
    nrow(Q3penalty_matches)
)

# Every shootout match must contribute one away-team observation.

stopifnot(
  sum(AwayTeamCounts$NumberOfMatches) ==
    nrow(Q3penalty_matches)
)

# Confirm that both final tables contain exactly five teams.

stopifnot(
  nrow(Top5HomeTeams) == 5
)

stopifnot(
  nrow(Top5AwayTeams) == 5
)

# Confirm that the final tables contain the required columns.

stopifnot(
  identical(
    names(Top5HomeTeams),
    c(
      "TeamName",
      "TeamCode",
      "NumberOfMatches"
    )
  )
)

stopifnot(
  identical(
    names(Top5AwayTeams),
    c(
      "TeamName",
      "TeamCode",
      "NumberOfMatches"
    )
  )
)

cat("All Question 3 tests passed successfully.")
```

# Question 4: Factors Associated with Match Outcomes

## Rationale

Stage and ExtraTime are categorical factors so these have been selected. available directly in 'Matches' and are more likely to be related to how a match is resolved.
The function table() produces cross tabulations of the data between each factor and Result. Raw Percentages are more useful than counts, for checking sample sizes. few players.Too few players in stages and extra time groups to be suitable for comparison. These are the number of matches, which are calculated by the function prop.table(..., margin = 1) The percentage for each factor category for the outcome.
One-hundred-percent stacked bar charts are used to show the Relative proportions of home team wins, away team wins and draws, and ignoring all groups with the same height, using a different order in the second coordinate. 'coord_flip()' makes all groups easier to read by changing their order in the second coordinate. long stage labels.
The analysis is descriptive in nature, and not explanatory. The interpretation also takes into account the different sample sizes and the assumption that the variables 'Stage' and 'ExtraTime' are related to each other.

## Overall match outcomes

```{r q4-overall-result}
# Count the total number of matches for each result.

OverallResult <- table(Matches$Result)

OverallResult
```

There are 545 home-team wins, 240 away-team wins, and 179 draws. Home-team wins
are therefore the most common outcome in the complete dataset.

## Factor 1: Competition Stage

### Create a table of outcomes by Stage

```{r q4-stage-table}
# Count each match outcome within every competition stage.

StageResultCount <- table(
  Matches$Stage,
  Matches$Result
)

# Convert the counts into row percentages.
# margin = 1 calculates percentages within each stage.

StageResultPercent <- round(
  prop.table(
    StageResultCount,
    margin = 1
  ) * 100,
  1
)

# Present the percentage table.

knitr::kable(
  StageResultPercent,
  caption = "Percentage of Match Outcomes by Competition Stage"
)
```

Each row represents one competition stage and adds to approximately 100%.

### Visualise Result by Stage

```{r q4-stage-chart, fig.width=9, fig.height=6}
ggplot(
  Matches,
  aes(
    x = Stage,
    fill = Result
  )
) +
  geom_bar(
    position = "fill",
    colour = "black"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  labs(
    title = "Match Outcomes by Competition Stage",
    x = "Competition stage",
    y = "Percentage of matches",
    fill = "Match result"
  ) +
  theme_minimal()
```

### Interpretation of Stage

There is a clear association between 'Result' and competition stage. During the group stage, 51.2% of games are won by the home side, 24.3% are won by the away side. That results in a relatively even distribution of wins and draws with 24.6% of them being draws. distribution.
Drops become less frequent in the knockout stages. They make up just 2.1% of round of 16 games and 2.9% of quarter finals, and no games in semi-finals, and Third place matches, or finals. This mirrors the knockout rules of which are: Typically have a winner.
76.3% of round of 16 games, 73.7% of semi-finals and 88.8% of finals are won by home teams.The home team wins 76.3% of round of 16 games, 73.7% of semi-finals and 88.8% of finals. and 81.0% of finals. These percentages are based on a much smaller, however. The 676 group stage games were won by just as many sides as the 125 samples. The data set includes, for instance, only 21 finals.
The second group stage is a more even one with 47.2% home victories and 30.6% away victories, and 22.2% draws. Overall, the strongest association of 'Stage' is with the frequency This could be due to the tournament format or team difference, but it's likely due to draws. Extracurricular activities based on the information contained within the text.


## Factor 2: Extra Time

### Create a table of outcomes by ExtraTime

```{r q4-extra-time-table}
# Count each match outcome according to whether extra time occurred.

ExtraTimeResultCount <- table(
  Matches$ExtraTime,
  Matches$Result
)

# Convert the counts into percentages within each ExtraTime group.

ExtraTimeResultPercent <- round(
  prop.table(
    ExtraTimeResultCount,
    margin = 1
  ) * 100,
  1
)

# Present the percentage table.

knitr::kable(
  ExtraTimeResultPercent,
  caption = "Percentage of Match Outcomes by Extra-Time Status"
)
```

In this variable:

```text
0 = no extra time
1 = extra time occurred
```

### Visualise Result by ExtraTime

```{r q4-extra-time-chart, fig.width=8, fig.height=5}
ggplot(
  Matches,
  aes(
    x = ExtraTime,
    fill = Result
  )
) +
  geom_bar(
    position = "fill",
    colour = "black"
  ) +
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  labs(
    title = "Match Outcomes by Extra-Time Status",
    x = "Extra time: 0 = No, 1 = Yes",
    y = "Percentage of matches",
    fill = "Match result"
  ) +
  theme_minimal()
```

### Interpretation of ExtraTime

The results of the matches also vary based on the 'ExtraTime' property. Among 891 matches without 24.1% in an away-team win, while in extra time, 56.5% of the games were won by the home team. 19.4% in a draw.
In 73 extra-time games, 57.5% had been won by the home team, 34.2% had been won by the away team. 8.2% of the matches ended in a draw. However, the proportions for the home-wins are nearly the same, Extra time on the scoreboard makes away-team victories about ten percentage points more likely. Draws are approximately eleven per cent less common, and occurs when it doesn't.
The lower draw proportion is understandable, due to the fact that additional time usually occurs. In the knockout rounds if there is a winner required. Some matches might still contain a 'Result' It was a draw and the advancing side was chosen by a penalty shootout at the end of the match. team.
As the additional group is quite small, percentage in additional group are more. Has been receptive to individual observations. The pattern thus indicates a association does not lead to any direct change of which side is playing, but just that they are playing together. wins.

## Overall discussion
Both of the two factors that have been selected, the factor 'Stage' produces the clearer and more The variations in the 'Result' field are interpretable and, in particular, the difference between the 'group' and 'group2' values is interpretable. matches, with or without draw, and knockout matches. uncommon.
A few variables are associated with ExtraTime, such as less drawing and more win away from home. The home-win proportion is stable, and the proportion of home wins drops. The factors are not independent as extra time is predominantly at the end, in the knockout phase. Therefore, An extra time part of the 'ExtraTime' pattern might not be due to additional time, but due to stage rules itself.
Other factors which are not measured, such as team quality, opponent power, tournament, etc, also play a role. The results, 'Result', may also be influenced by year, and competition format. The analysis should should therefore be displayed as patterns and relationships within the data set, not as an evidence for a specific result of a match.


## Code testing

```{r q4-code-testing}
# Check that Result contains three categories.

stopifnot(
  nlevels(Matches$Result) == 3
)

# Check that Stage and ExtraTime are factors.

stopifnot(
  is.factor(Matches$Stage)
)

stopifnot(
  is.factor(Matches$ExtraTime)
)

# Check that every match was included in the Stage table.

stopifnot(
  sum(StageResultCount) == nrow(Matches)
)

# Check that every match was included in the ExtraTime table.

stopifnot(
  sum(ExtraTimeResultCount) == nrow(Matches)
)

# Check that the percentage-table rows total approximately 100%.

stopifnot(
  all(
    abs(
      rowSums(StageResultPercent) - 100
    ) <= 0.2
  )
)

stopifnot(
  all(
    abs(
      rowSums(ExtraTimeResultPercent) - 100
    ) <= 0.2
  )
)

cat("All Question 4 tests passed successfully.")
```
#Question 5: GitHub repository and commit on this platform

## GitHub Repository

The complete source code, datasets, report files, and development history for
this assignment are available in the following GitHub repository:

<https://github.com/mapphat2005-svg/COMP1013--Analysis-Programming-assignment>

## Repository Structure
- data/ contains all task dataset 

- scripts/ contains the R scripts used to analyze the question 1 to 4

- report/ contains the official report

- figures/ contains the image used in the report

## Git Commit History

The following screenshot shows the Git commit history created during the
development of the assignment.

```{r task5-git-history, echo=FALSE, out.width="100%", fig.align="center"}
knitr::include_graphics(
  here::here(
    "output",
    "figures",
    "git_commit_history.png"
  )
)
```

## Explanation of Development Commits

The assignment was developed through a series of separate commits rather than
being uploaded as one final version. Each commit represents a meaningful stage
of development.

1. **Initial commit**  
   This commit created the initial GitHub repository and established the main
   branch.

2. **Create GitHub repository and upload datasets for this assignment**  
   This commit added the four original CSV datasets and created the folders for
   data, scripts, report files, figures, and tables.

3. **Complete Question 1 data cleaning and penalty visualisation**  
   This commit added the code for inspecting the datasets, replacing question
   marks with missing values, converting variable types, processing penalty
   values, and visualising the distribution of `HomePenalty`.

4. **Complete Question 3 penalty shootout team rankings**  
   This commit added the analysis that identified the five home teams and five
   away teams most frequently involved in penalty shootouts.

5. **All R scripts were used in the assignment**  
   This commit reorganised the code from the different questions into one
   complete R script. The consolidation made it possible to run the assignment
   analysis from a single file.

7. **Finalise report and submission files**  
   This final commit added the completed PDF report, the Git commit-history
   screenshot, and the final formatting and proofreading changes.
